---
title: "ggplot2 demo, part 1"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading packages

```{r}
library(readr)
library(dplyr)
library(ggplot2)
```

## Illustrative example

### Inspecting the `mpg` dataset

The `mpg` dataset provides information about the fuel economy of popular car models in 1999 and 2008. It comes included with ggplot2:

```{r}
glimpse(mpg)
```

### Basic scatter plot

We'll plot the highway mile per gallon (`hwy`) against the engine displacement in liter (`displ`):

```{r}
ggplot(data=mpg, mapping=aes(x=displ, y=hwy)) + 
  geom_point()
```

### Titles, axes, and position scales

Adding title and axes labels to a plot:

```{r}
ggplot(data=mpg, mapping=aes(x=displ, y=hwy)) + 
  geom_point() + 
  ggtitle("Highway fuel efficiency vs engine size") +
  xlab("Engine displacement (L)") + 
  ylab("highway mile per gallon")
```

### Customizing theme

Use `theme_bw()` instead the default `theme_gray()`, and change the font size of a few elements via `theme()`

```{r}
ggplot(data=mpg, mapping=aes(x=displ, y=hwy)) + 
  geom_point() + 
  ggtitle("Highway fuel efficiency vs engine size") +
  xlab("Engine displacement (L)") + 
  ylab("highway mile per gallon") +
  theme_bw() + theme(
    plot.title = element_text(face="bold", size=20),
    axis.title = element_text(size=16),
    axis.text.x = element_text(size=14),
    axis.text.y = element_text(size=14)
  )
```

### Assigning parts of the plot composition to variables

Assigning part of the plot composition to variables

```{r}
mpg_points <- ggplot(data=mpg, mapping=aes(x=displ, y=hwy)) + 
  geom_point()

titles <- labs(
  title="Highway fuel efficiency vs engine size",
  x = "Engine displacement (L)",
  y = "highway mile per gallon"
)

my_theme <-  theme_bw() + theme(
    plot.title = element_text(face="bold", size=20),
    axis.title = element_text(size=16),
    axis.text.x = element_text(size=14),
    axis.text.y = element_text(size=14)
  )
```

Compose and print the plot from variables

```{r}
mpg_points + titles + my_theme
```

Export the resulting plot as svg:

```{r}
ggsave("mpg_scatter.svg", width=6, height=4)
```

### Adding more aesthetics mapping

Adding color and shape to distinguish between different drive mode

```{r}
ggplot(mpg, aes(x=displ, y=hwy, color=drv, shape=drv)) + 
  geom_point() + titles + my_theme
```

Plot highway mpg (`hwy`) against city mpg (`cty`), with size indicating engine displacement:

```{r}
ggplot(mpg, aes(x=cty, y=hwy, size=displ)) + geom_point() + 
  ggtitle("Two measures of fuel efficiency") + 
  xlab("city mile per gallon") +
  ylab("highway mile per gallon") +
  my_theme
```

Customize visual properties independent of attribute values of observations

```{r}
ggplot(mpg, aes(x=displ, y=hwy)) + 
  geom_point(color="blue")
```

### Adding additional layers

Adding an additional layer (linear least square fit). Note that BOTH the fit and the data are segregated by drive type and use the same underlying data

```{r}
ggplot(mpg, aes(x=displ, y=hwy, color=drv, shape=drv)) + 
  geom_point() +
  geom_smooth(method="lm")
```

If we want to have a common best fit line across drive type, we need to restrict the color and shape mapping to the scatter plot:

```{r}
ggplot(mpg, aes(x=displ, y=hwy)) + 
  geom_point(aes(color=drv, shape=drv)) +
  geom_smooth(method="lm", color="black")
```

Figure with data from two sources: plot the `mpg` observations (1999-2008) against a rolling fit of data from `mtcars` observations (1973-1974)

```{r}
ggplot() + 
  geom_point(aes(x=displ, y=hwy, color=drv, shape=drv), mpg) +
  geom_smooth(aes(x=0.0163871 * disp, y=mpg), mtcars, method="loess", color="black")
```

