---
title: "ggplot2 demo, part 2"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading packages

```{r}
library(readr)
library(dplyr)
library(ggplot2)
```

## Plot types (geoms)

### Line plot: `geom_line()` and `geom_path()`

Use the `economics` dataset included in ggplot2

```{r}
economics
```

Plot the percent saving rate (`psavert`) as a function of date (`date`)

```{r}
ggplot(economics, aes(x=date, y=psavert)) + geom_line()
```

Plot progression of unemployment population (`unemploy`) and unemployment time (`uempmed`) as path

```{r}
ggplot(economics) +
    geom_path(aes(x = unemploy, y = uempmed, color=date))
```

### Including uncertainty: `geom_errorbar()` and `geom_ribbon()`

Read the monthly weather normal (based on statistics from 1981 to 2010) at the Portege Bay NOAA station:

```{r}
monthly_normal <- read_csv("data/monthly_normal.csv")
```

```{r}
monthly_normal
```


Plot the average monthly temperature, together with error bars

```{r}
ggplot(monthly_normal, aes(x=month, y=Tavg_normal)) +
  geom_point() +
  geom_errorbar(aes(
    ymin = Tavg_normal - Tavg_stddev, 
    ymax = Tavg_normal + Tavg_stddev
  ), width=0.5)
```

Plot average monthly temperature, together with error ribbon

```{r}
ggplot(monthly_normal, aes(x=month, y=Tavg_normal)) +
  geom_point(size=2) + geom_line() +
  geom_ribbon(aes(
    ymin = Tavg_normal - Tavg_stddev, 
    ymax = Tavg_normal + Tavg_stddev),
    alpha=0.1, color="blue", fill="cyan"
  )
```

### Bar plot: `geom_bar()` and `geom_col()`

Bar plot counting number of samples from each class of cars in `mpg`

```{r}
ggplot(mpg, aes(x=class)) + geom_bar()
```

Summarize average highway mile per gallon for each class

```{r}
mpg_by_class <- mpg |> group_by(class) |> 
  summarize(hwy=mean(hwy), cty=mean(cty))
```

Plot class against (averaged) highway mile per gallon:

```{r}
ggplot(mpg_by_class, aes(x=class, y=hwy)) + geom_col()
```

Horizontal columns instead of vertical ones

```{r}
ggplot(mpg_by_class, aes(y=class, x=hwy)) + geom_col()
```

Use color to indicate different drive mode within a class

```{r}
ggplot(mpg, aes(x=class)) + geom_bar(aes(fill=drv))
```

### Histogram `geom_histogram()` and `geom_density()`

Histogram of highway mile per gallon

```{r}
ggplot(mpg, aes(x=hwy)) + geom_histogram()
```

Specify precise bins with `binwidth` and `boundary`

```{r}
ggplot(mpg, aes(x=hwy)) + geom_histogram(binwidth=2.5, boundary=10)
```

Separate different drive mode using colors. Note that by default the different groups stack on top of each others

```{r}
ggplot(mpg, aes(x=hwy)) + 
  geom_histogram(aes(fill=drv), binwidth=2.5, boundary=10)
```

### Categorical variables: `geom_boxplot()`

Box plot of highway mile per gallon against class

```{r}
ggplot(mpg, aes(x=class, y=hwy)) + geom_boxplot()
```

Box plot including both class and drive mode

```{r}
ggplot(mpg, aes(x=class)) + 
  geom_boxplot(
    aes(y=hwy, color=drv), 
    position=position_dodge2(preserve="single")
  )
```

Horizontal box plot

```{r}
ggplot(mpg, aes(y=class, x=hwy)) + geom_boxplot()
```

### 3-dimensional data: `geom_contour()` and friends

Loading data (gridded data formatted as a dataframe)

```{r}
dframe <- read_csv("data/polynomial_mul_gaussian.csv")
```

```{r}
dframe
```

Create a filled contour plot based on the value of z

```{r}
ggplot(dframe, aes(x=x, y=y, z=z)) + geom_contour_filled()
```

Create a contour plot instead of filled contour plot

```{r}
ggplot(df, aes(x=x, y=y, z=z)) + 
  geom_contour(aes(color=after_stat(level)))
```

Setting finer contour levels

```{r}
ggplot(df, aes(x=x, y=y, z=z)) + 
  geom_contour(aes(color=after_stat(level)), binwidth=0.05)
```

### 3-dimensional data: `geom_tile()` and `geom_raster()`

Instead of contour plot, use tile of values

```{r}
ggplot(df, aes(x=x, y=y, fill=z)) + geom_tile()
```

More efficient version using `geom_raster()` rather than `geom_tile()`

```{r}
ggplot(df, aes(x=x, y=y, fill=z)) + geom_raster()
```

Turn on interpolation:

```{r}
ggplot(df, aes(x=x, y=y, fill=z)) + geom_raster(interpolate=TRUE)
```

### Annotations

Line annotations:

```{r}
ggplot(monthly_normal, aes(x=month, y=Tavg_normal)) +
  geom_hline(yintercept=53.5, linewidth=2, color="orange") + 
  geom_point() +
  geom_errorbar(aes(
    ymin = Tavg_normal - Tavg_stddev, 
    ymax = Tavg_normal + Tavg_stddev
  ), width=0.5)
```

```{r}
ggplot(mpg_by_class, aes(y=class, x=hwy)) + 
  geom_col() +
  geom_vline(xintercept = mean(mpg$hwy), color="cyan", linewidth=2)
```
