---
title: "ggplot2 demo, part 3"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading packages

```{r}
library(stringr)
library(ggplot2)
library(dplyr)
```

## Titles, axes, and position scales

### Titles and axes labels

Adding x- and y-axis labels and title to a plot:

```{r}
mpg_point <- ggplot(mpg, aes(x=displ, y=hwy)) + geom_point()
  
mpg_point + ggtitle("Highway fuel efficiency vs engine size") +
  xlab("Engine displacement (L)") + ylab("highway mile per gallon")
```

Alternative implementation using `labs()`

```{r}
mpg_point + labs(
  title="Highway fuel efficiency vs engine size",
  x = "Engine displacement (L)",
  y = "highway mile per gallon"
)
```

Alternative implementation using the underlying `scale...()` functions

```{r}
mpg_point + ggtitle("Highway fuel efficiency vs engine size") +
  scale_x_continuous(name="Engine displacement (L)") + 
  scale_y_continuous(name="highway mile per gallon")
```
`
A rich-text y label created with `plotmath`, and unicode character shown in title

```{r}
ggplot(mpg, aes(x=displ, y=hwy/cty)) + 
  geom_point(size=2) +
  labs(
    title="Highway fuel efficiency ε vs engine size",
    x = "Engine displacement (L)",
    y = expression(bgroup( "(", frac("hwy","cty"), ")" )) 
  )
```

### Customizing limits of plot

Plotting the saving rate of the `economics` dataset using defaults:

```{r}
ec_plot <- ggplot(data=economics, aes(x=date, y=psavert)) + geom_line()

print(ec_plot)
```

**Subset** the data to go from 1990 to 2010 using `xlim()` (note that ggplot automatically insert some buffering space between the left and right limits):

```{r}
ec_plot + xlim(as.Date(c("1990-01-01", "2010-12-31")))
```

**Zoom in** to the data to go from 1990 to 2010 using `xlim()`:

```{r}
ec_plot + coord_cartesian(xlim=as.Date(c("1990-01-01", "2010-12-31")))
```

### Customizing ticks locations and labels

Plot of mpg data using the defaults:

```{r}
mpg_plot <- ggplot(mpg, aes(x=displ, y=hwy)) + geom_point()

print(mpg_plot)
```

Use `scale_x_continuous()` to set the location of ticks:

```{r}
mpg_plot + scale_x_continuous(limits=c(0,8), breaks=c(0,2,4,6,8))
```

Use `scale_x_continuous()` to customize the x labels:

```{r}
mpg_plot + scale_x_continuous(
  breaks = seq(2,7), 
  labels = str_c(seq(2,7), "L")
)
```

### Transforming axes scales

Plotting simple exponential data

```{r}
exp_data <- data.frame(
  x = seq(0, 10, length.out=101),
  y = exp(-seq(0, 10, length.out=101))
)

exp_plot <- ggplot(exp_data, aes(x=x, y=y)) + geom_line()

print(exp_plot)
```

Plotting the same data in log-linear scale

```{r}
exp_plot + scale_y_continuous(transform="log10")
```

Same plot, with friendier y=axis labels (again using `plotmath`)

```{r}
exp_plot + scale_y_continuous(
  transform="log10",
  breaks=10^seq(0,-4, -1),
  labels=expression("10"^"0", "10"^"-1", "10"^"-2", "10"^"-3", "10"^"-4")
)
  
```

### Setting aspect ratio

Fix the aspect ration to be 1:1

```{r}
pow_data <- data.frame(
  x = seq(0, 1, length.out=101),
  y = seq(0, 1, length.out=101) ^ 2
)

pow_base <- ggplot(exp_data, aes(x=x, y=y)) + geom_line()

pow_base + coord_fixed()
```

Change the aspect ratio to 1:0.5

```{r}
pow_base + coord_fixed(0.5)
```

### Second axis

Check summary statistics to decide how to align the the axes

```{r}
summary(economics)
```

Decision: unemploy with limits c(0, 18000), corresponding to uempmed limits of c(0, 10)

Plotting unemployment number and unemployment duration on the same figure. 

Note also the use of "constant" aesthetics

```{r}
ggplot(economics) + 
  geom_line(aes(x=date, y=unemploy, linetype="unemploy", color="unemploy")) + 
  geom_line(aes(x=date, y=1800 * uempmed, linetype="uempmed", color="uempmed")) +
  scale_y_continuous(
    limits=c(0, 18000), 
    sec.axis = sec_axis(~ . / 1800, name="uempmed")
  ) + labs(
    color="data series", linetype="data series"
  )
```

Use the convenient function defined in `sec_axis_pair()`

```{r}
source("sec_axis_pair.R")

pair <- sec_axis_pair(c(0, 18000), c(0, 10))

ggplot(economics) + 
  geom_line(aes(x=date, y=unemploy, linetype="unemploy", color="unemploy")) + 
  geom_line(aes(x=date, y=pair$func(uempmed), linetype="uempmed", color="uempmed")) +
  scale_y_continuous(
    limits=c(0, 18000), 
    sec.axis = sec_axis(pair$formula, name="uempmed")
  ) + labs(
    color="data series", linetype="data series"
  )
```

### Date and datetime axes

Default setting for date axis

```{r}
ec_plot <- ggplot(economics, aes(x=date, y=psavert)) + geom_line()
print(ec_plot)
```

Make ticks every 5 years

```{r}
ec_plot + scale_x_date(date_breaks = "5 years", date_labels="%Y")
```

Change formatting of date label

```{r}
ec_plot + scale_x_date(date_breaks = "10 years", date_labels="%Y-%m-%d")
```

## Scales for other aesthetics

### Scales for edge and fill colors

Data behind raster plots

```{r}
fine_frame <- read_csv("data/polynomial_mul_gaussian_finer.csv")
```

Customize raster fills (a _continuous_ color scale) using the viridis color set (see <https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html> for documentation)

```{r}
cont_filled <- ggplot(fine_frame, aes(x=x, y=y, fill=z)) + geom_raster()

cont_filled + scale_fill_viridis_c(option = "magma")
```

Repeat the above but use the brewer color set (see <https://r-graph-gallery.com/38-rcolorbrewers-palettes.html> for illustration)

```{r}
cont_filled + scale_fill_distiller(palette = "RdYlBu")
```

Customize fill color of columns in a `geom_col()` plot (a _discrete_ color scale) using the viridis color set:

```{r}
mpg_by_class <- mpg |> group_by(class) |> 
  summarize(hwy=mean(hwy), cty=mean(cty))

mpg_class_plot <- ggplot(mpg_by_class, aes(x=class, y=hwy, fill=class)) + 
  geom_col()

mpg_class_plot + scale_fill_viridis_d(option="turbo")
```

Repeat the same using the brewer scale

```{r}
mpg_class_plot + scale_fill_brewer(palette="Dark2")
```

### Linear versus logarithmic color scale

Load data that represent an exponentiated parabolic saddle point

```{r}
exp_saddle <- read_csv("data/exponential_saddle.csv")
```

Plot and exponentiated saddle point in linear color scale

```{r}
exp_saddle_plot <- ggplot(df2, aes(x=x, y=y, fill=z)) + geom_raster()
exp_saddle_plot + scale_fill_viridis_c(option = "magma")
```

Plot the same in logrithmatic color scale

```{r}
exp_saddle_plot + scale_fill_viridis_c(option = "magma", transform="log10")
```

### Scales for shape and linetype

Specifying shape using `scale_shape_manual()`:

```{r}
ggplot(mpg, aes(x=displ, y=hwy, shape=drv)) + geom_point() + 
  scale_shape_manual(
    values=c("square open", "diamond open", "triangle open")
  )
```

Specify linetype using `scale_linetype_manual()`

```{r}
ggplot(mutate(economics, unemprt=100 * unemploy / pop), aes(x=date)) + 
  geom_line(aes(y=psavert, linetype="saving")) +
  geom_line(aes(y=unemprt, linetype="unemployment")) +
  labs(linewidth="rate", y="rate") +
  scale_linetype_manual(
    values=c("dashed","dotted")
  ) 
```

For both cases, see <https://ggplot2.tidyverse.org/articles/ggplot2-specs.html> for documentation

### Legend and guides

Customize the legend using `name`, `breaks`, and `labels` arguments of `scale...()`

```{r}
ggplot(mpg, aes(x=displ, y=hwy, shape=drv)) + geom_point() + 
  scale_shape_manual(
    values=c("square filled", "diamond filled", "triangle filled"),
    breaks=c("4", "r", "f"), 
    labels=c("4-wheel drive", "rear-wheel drive", "forward-wheel drive"),
    name="drive mode"
  )
```

